Retrofitting Effect Handlers onto OCaml

### 2021-03-11 retro-concurrency.tex
updates

\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\author{KC Sivaramakrishnan, Stephen Dolan, Leo White, Tom Kelly, Sadiq Jaffer and Anil Madhavapeddy}
\acmConference[Draft Article]{}{21 Nov 2020}{Under Blind Review}%

@@

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

\subsection{An effectful calculator}

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, coroutines, generators and asynchronous I/O to be
composably expressed. For example, consider the task of handling
division-by-zero in a calculator application. Different calculators handle
division-by-zero differently. In the iOS 13 Calculator (|iOS|),
division-by-zero returns |Error|. In the Google search calculator (|gsearch|)
|0/0| returns |Error|, but for |n!=0| the |n/0| returns |Infinity| with the
same sign as |n|. How do we capture all these different behaviours without
changing the evaluator?

Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision.
Conditions provide a \emph{restart} mechanism for non-fatal errors to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We now implement
this using effect handlers in OCaml syntax:

\begin{lstlisting}[numbers=left]
type sign = P | N
type val = Z of int | Error | Inf of sign
type exp = Val of val | ... | Div of exp*exp
effect DivBy0 : int -> val
let rec eval e =
  match e with
  | Val v -> v
  | ...
  | Div (e1,e2) ->
    match eval e1, eval e2 with
    | Z v1, Z v2 ->
        if v2 = 0 then perform (DivBy0 v1)
        else Z (v1/v2)
    | Error, _ -> Error | Inf s, _ -> Inf s
    | ...
let (/) n d = Div (Val (Z n), Val (Z d))
\end{lstlisting}

Line 4 declares an effect |DivBy0| that is parameterised by an integer, which
when performed returns the result of division-by-zero. If the denominator of
the division is |0|, then we perform the |DivBy0| effect with the numerator
value (line 12) so that the context may decide the return value. We obtain the
|iOS| semantics by wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v (* value case *)
| effect (DivBy0 _) k -> continue k Error (* effect case *)
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).

\subsection{Asynchronous I/O}
\label{sec:aio}

We can modularly extend the calculator to read user input using the |Read|
effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string (perform (Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
		register_async_input_line (chan, k);
		continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither the division-by-zero handler nor the I/O handler needs to be aware
of the other.

@@

\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program
does not break under OCaml extended with effect handlers. OCaml code that does
not use effect handlers will pay negligible performance and memory cost.

\noindent \textbf{R2: Tool compatibility.} OCaml programs with effect handlers
produce well-formed backtraces and remain compatible with program analysis
tools such as debuggers and profilers that inspect the stack using DWARF unwind
tables.

\noindent \textbf{R3: Effect handler efficiency.} The program must accommodate
millions of continuations at the same time to support highly-concurrent
applications. Installing effect handlers, capturing and resuming continuations
must be fast.

\noindent \textbf{R4: Forwards compatibility.} As a cornerstone of modularity,
we also want blocking I/O code to \emph{transparently} be made asynchronous
with the help of effect handlers.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, and the |monad| version was 8.69$\times$ slower than |cps|.

\textbf{\textit{Chameneos.}} Chameneos~\cite{Chameneos} is a concurrency game
aimed at measuring context switching and synchronization overheads. Our
implementation uses |MVars| for synchronization. We compare effect handler
(|MC|), concurrency monad (|monad|) and Lwt, a widely used concurrency
programming library for OCaml (|lwt|) versions. We observed that |MC| was the
fastest, and |monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library that spawns a thread per
request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
style and does not have the notion of a thread per request. For comparison, we
include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=0.9\linewidth]{http-benchmarks/throughput.pdf}
		%\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=0.9\linewidth]{http-benchmarks/latency.pdf}
		%\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	%\vspace{-3mm}
	\caption{Web server performance.}
	%\vspace{-5mm}
\end{figure}

>>>>

\documentclass[sigplan,screen]{acmart}
\setcopyright{none}
\acmPrice{}
\acmDOI{10.1145/3453483.3454039}
\acmYear{2021}
\copyrightyear{2021}
\acmSubmissionID{pldi21main-p82-p}
\acmISBN{978-1-4503-8391-2/21/06}
\acmConference[PLDI '21]{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 20--25, 2021}{Virtual, UK}
\acmBooktitle{Proceedings of the 42nd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '21), June 20--25, 2021, Virtual, UK}

@@

\author{KC Sivaramakrishnan}
\affiliation{
  \institution{IIT Madras}            %% \institution is required
	\country{India}
}
\email{kcsrk@cse.iitm.ac.in}          %% \email is recommended

\author{Stephen Dolan}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{stephen.dolan@cl.cam.ac.uk}

\author{Leo White}
\affiliation{
  \institution{Jane Street}            %% \institution is required
	\country{UK}
}
\email{leo@lpw25.net}

\author{Sadiq Jaffer}
\affiliation{
  \institution{Opsian and OCaml Labs} %% \institution is required
	\country{UK}
}
\email{sadiq@toao.com}

\author{Tom Kelly}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{tom.kelly@cantab.net}

\author{Anil Madhavapeddy}
\affiliation{
  \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{avsm2@cl.cam.ac.uk}          %% \email is recommended

@@

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
handle |Read| differently. For example, |Read| may be implemented by performing
a blocking read on the input channel or performing the read asynchronously by
offloading it to an event loop such as |libuv|. Thanks to the separation of
effectful operations from their interpretations, effect handlers enable new
approaches to modular programming.

\subsection{Asynchronous I/O}

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. Let us illustrate the power of
effect handlers in OCaml by implementing a simple lightweight thread library
with support for forking new threads and yielding control to other threads. We
will then extend this scheduler with support for synchronously reading from
channels and subsequently make it asynchronous without changing the client code
for asynchrony.

Many programming languages, including OCaml, require non-local change to the
code in order to support asynchronous I/O, often leading to a dichotomy
synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
primitive support for concurrency, users will have to use monadic I/O libraries
such as Lwt~\cite{lwt} and Async~\cite{async}. Languages such as GHC Haskell
and Go provide lightweight threads but bake their implementation into the
runtime system. With effect handlers, asynchronous I/O can be implemented
directly in OCaml.

In order to support forking and yielding threads, we declare effects the
following effects:

\begin{lstlisting}
effect Fork : (unit -> unit) -> unit
effect Yield : unit
\end{lstlisting}

The |Fork| effect takes a thunk which is spawned as a concurrent thread, and
|Yield| yields control to another thread in the scheduler queue. We can define
helper functions to perform these effects:

\begin{lstlisting}
let fork f = perform (Fork f)
let yield () = perform Yield
\end{lstlisting}

The implementation of the scheduler queue is defined in the |run| function,
which handles the effects appropriately:

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push k runq in
  let rec run_next () =
    match Queue.take_opt runq with
    | None -> ()
    | Some k -> continue k ()
  in
  let rec spawn f =
    match f () with
		| () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
  in
  spawn main
\end{lstlisting}

The function |spawn| (line 9) evaluates the computation |f| in an effect
handler. The computation |f| may return normally with a value, or perform
effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
|Yield| and binds |k| to the continuation of the corresponding |perform|
delimited by this handler. The scheduler queue |runq| maintains a queue of
these continuations. |suspend| pushes continuations into the queue, |run_next|
pops continuations from the queue and resumes them with |()| value using the
|continue| primitive. In the case of |Yield| effect, we suspend the current
continuation |k| and resume the next available continuation. In the case of
|Fork f'| effect, we suspend the current continuation and recursively call
|spawn| on |f'| in order to run |f'| concurrently.

Effect handlers provide modularity. Observe that we can change the scheduling
algorithm from FIFO to LIFO by changing the scheduler queue to a stack. We can
implement support for synchronous read from channels by adding the following
case to the effect handler in |spawn|:

\begin{lstlisting}
let rec spawn f =
  match f () with
	...
  | effect (Read ic) k -> continue k (input_line ic)
\end{lstlisting}
%
which uses OCaml standard library's |input_line| function to read a line from
the channel |ic| and resume the continuation |k| with the resultant string.
However, performing reads synchronously blocks the entire scheduler, preventing
other threads to run until the I/O is completed.

We can make the I/O asynchronous by modifying the |run| function as follows:

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
	let suspend k =
	  Queue.push (fun () -> continue k ()) runq
	in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.take_opt runq with
    | None ->
      begin match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let (done_,todo) = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) done_;
        pending_reads := todo;
        run_next ()
      end
    | Some f -> f ()
  in
  let rec spawn f =
    match f () with
		...
    | effect (Read ic) k ->
        pending_reads := (ic,k)::!pending_reads;
        run_next ()
  in
  ...
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 6). We handle the |Read| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 25), allowing
other threads in the scheduler to run.

When the scheduler queue is empty, the |run_next| function performs the pending
reads. We abstract away the details of the event-based I/O using the |do_reads|
function (line 13). |do_reads| takes a list of pending reads and blocks until
at least one of the reads succeeds. It returns a pair of lists |done_| and
|todo|. |done_| contains the result strings from successful reads and
corresponding continuations. These continuations are arranged to be resumed
with the read result and pushed into the scheduler queue. |todo| contains the
channels on which input is still pending and their corresponding continuations.
|pending_reads| is updated to point to the |todo| list so that they may be
attempted later. Observe that all of the changes to add asynchrony are
localised to the |run| function, and the computation that performs these
effects can remain in \emph{direct-style} (as opposed to the monadic-style in
Lwt and Async).

@@

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} A well-behaved OCaml program does
		not break under OCaml extended with effect handlers. OCaml code that does
		not use effect handlers will pay negligible performance and memory cost.

	\item \textbf{Tool compatibility.} OCaml programs with effect handlers
		produce well-formed backtraces and remain compatible with program analysis
		tools such as debuggers and profilers that inspect the stack using DWARF
		unwind tables.

	\item \textbf{Effect handler efficiency.} The program must accommodate
		millions of continuations at the same time to support highly-concurrent
		applications. Installing effect handlers, capturing and resuming
		continuations must be fast.

	\item \textbf{Forwards compatibility.} As a cornerstone of modularity, we
		also want blocking I/O code to \emph{transparently} be made asynchronous
		with the help of effect handlers.
\end{enumerate}

@@

\subsubsection{Generators}

Generators allow data structures to be traversed on demand. Many languages
including JavaScript and Python provide generators as a built-in primitive.
Using effect handlers (|MC|), given any data structure (|'a t|) and its
iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
generator function (|val next : unit -> 'a option|). The code is available in
the supplementary material. We evaluate the performance of traversing a
complete binary tree of depth 25 using this generator. This involves $2^{26}$
stack switches in total. For comparison, we implemented a hand-written,
selective CPSed~\cite{Nielson01}, defunctionalised~\cite{Danvy01} version
(|cps|) and a concurrency monad (|monad|) version of the generator for the
tree. Both |cps| and |monad| versions are specialised to the binary tree with
the usual caveats of not using the stack for function calls. We observed that
the |cps| version was the fastest, thanks to specialisation and hand
optimisation. |MC| version was only 2.76$\times$ slower than |cps| while being
a generic solution, and the |monad| version was 8.69$\times$ slower than |cps|.

\subsubsection{Chameneos}

Chameneos~\cite{Chameneos} is a concurrency game aimed at measuring context
switching and synchronization overheads. Our implementation uses |MVars| for
synchronization. We compare effect handler (|MC|), concurrency monad (|monad|)
and Lwt, a widely used concurrency programming library for OCaml (|lwt|)
versions. We observed that |MC| was the fastest, and |monad| (|lwt|) was
1.67$\times$ (4.29$\times$) slower than |MC|.

\subsubsection{Webserver}

As one of our primary goals of adding effect handlers to OCaml is to support
direct-style concurrency, we implement a full-fledged HTTP/1.1 web server using
effect handlers (|MC|). We use effect handlers to implement a user-level
threading library that spawns a thread per request. We use
\texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in
monadic style and does not have the notion of a thread per request. For
comparison, we include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

\begin{figure}
	\begin{minipage}{0.49\linewidth}
		\centering
		\includegraphics[width=\linewidth]{http-benchmarks/throughput.pdf}
		%\vspace{-2mm}
		\subcaption{Throughput}
		\label{grf:throughput}
	\end{minipage}
	\begin{minipage}{0.49\linewidth}
		\includegraphics[width=\linewidth]{http-benchmarks/latency.pdf}
		%\vspace{-2mm}
		\subcaption{Tail latency}
		\label{grf:latency}
	\end{minipage}
	%\vspace{-3mm}
	\caption{Web server performance.}
	%\vspace{-5mm}
\end{figure}
